Depression and suicidality as costly signals of need

### 2021-04-12 Paper.rmd
Changed signaling intro and minor edits to discussion

Due to the ease in which signals can be faked [@Zahavi1975; @mokkonen2016] and the costs of accepting a deceptive signal as reliable [@dawkins1978; @johnstone1993; @Hasson1994], demonstrating that one’s need is genuine has likely been a recurrent adaptive problem since humans evolved the capacity to provide instrumental support to one another. One possible solution to this is signaling in ways that are unprofitable to fake due to the costs involved. [@Zahavi1975]. This is possible whenever the signal in question is comparatively cheaper to produce for those who have the relevant underlying quality relative to those who lack it [@grafen1990]. In these instances, individuals with the quality maximize their fitness at greater levels of investment in the signal compared to those who lack the quality or possess it to a lesser degree [@zahavi1977]. The end result of this is expected to be a signal that either tends to not be worth faking or is impossible to fake to a large degree due to selection against fakers constraining the physiology behind the signal [@zahavi1977; @gangestad2005].

<!-- some overlap with my wikipedia article -->

@@

Distinguishing between the information provided by signals versus cues is a major challenge for signaling approaches to behavior. A signal differs from a cue in that it is intentionally sent by an organism to alter the behavior or thoughts of other individuals, while a cue is any piece of information an organism uses to alter its behavior or physiological state that was created without the intent to do so [@MaynardSmithHarper1995]. Regarding the signals of interest, suicide attempts are more likely to take the form of pure, external signals than other symptoms of depression. This is because low mood's impact on behavior for non-signaling benefits likely conveys information as cues in addition to what might exist for signaling purposes. In other words, the signaling component of depression may come from the upregulation of symptoms beyond what would be optimal for non-signaling purposes rather than the creation of entirely new symptoms. Suicidality is likely an exception to this, due to the likelihood most suicide attempts likely do not have non-signaling functions [see @deCatanzaro1991 for a possible exception].

@@

Consistent with costly signaling theory, costlier signals resulted in greater perceptions of need and increased the likelihood of participants helping the signaler than the less costly signals. This effect was largest in the "brother-in-law" and "romantic partner" vignettes, both of which involved claims of assault against participants' imagined daughters. However, the effect of signal type was much smaller in both the "basketball coach" and the "thwarted marriage" vignette. In the "basketball coach" vignette, part of the reason for this may have been the lack of relatedness between participants and the alleged victim. However, this was still a surprising finding due to the relatively small size of the request (i.e., recommend the coach be suspended rather than kick out family members or a romantic partner). With the thwarted marriage vignette, the fact that support came at the cost of one's younger daughter may explain the relatively small effect of signal strength. However, data from the Indian sample appears to be of lower quality, limiting our confidence in this interpretation (see limitations for more information). It is also possible that our research in designing the vignettes was insufficient for the population or the content of the vignette.

In the "brother-in-law" and "romantic partner" scenarios, *Crying* had little effect on the magnitude of pro-signaler responses, while both *Depression* conditions increased support to similar degrees. One potential reason for this is the increase in symptoms from mild depression to depression was not costly enough to result in a large effect as the symptoms only differed in severity (e.g., grades dropping from As to Bs [mild depression] or Cs [depression]). Such small changes may be less impactful in vignette form than in real-life scenarios in which the effects of signaling may increase in severity and be prolonged over a lengthy period of time.

While the effect of *Suicide attempt* on belief at T2 was similar to the *Depression* conditions, it resulted in greater likelihood of support. This may be due to our decision to measure belief in need as the degree participants believed the claims of the signaler (e.g., if the assaults actually happened) rather than a more inclusive measure. For this reason, signal cost may have provided additional information about the costliness of the signaler's adversity in addition to information relevant to whether the event claimed actually happened.

Support for this interpretation comes from our mediation analyses, which showed that the likelihood of help was largely, but not entirely, mediated by signal's effect on belief in need in this study but totally mediated by belief in need when measured as belief in need of financial support in the pilot study when comparing both depression and suicide attempts (confirm suicide) to verbal requests. However, the partial mediation in this study may have also arisen if participants viewed the signal as providing information about adversity outside of the signaler's claims. For example, participants who saw the "brother-in-law" vignette may have thought the daughter was lying about the assault due to jealously over her cousin, with the stronger signals still conveying the severity of the competition between her and her cousin.

## Relevance to costly signaling

@@

The need to explain suicide's connection to depression is often overlooked by explanations for major depression, particularly those which propose non-signaling functions. One explanation for their connection, is that both suicidality and depression may function to limit negative fitness outcomes in times of substantial adversity [@andrews2006; @Hagen2002; @Hagen2003; @HagenEtAl2008; @nock2008; @Rosenthal1993; @Stengel1956; @SymeEtAl2016], with costlier signals being required to elicit support when the payoffs to faking need are high enough so that less costly signals are not reliable indicators of one's need [@Rosenthal1993]. For this hypothesis to be correct, instrumental support would need to be provided frequently enough with large enough impacts so that the costs of signaling are outweighed by the benefits on average. In support of this, we found exposure to depression and suicide attempts resulted in greater belief in need and increased likelihood of support in fictional scenarios. This coupled with findings that depression can simultaneously elicit both positive and negative responses [@HokansonEtAl1986; @stephens1987], suggests that more attention should be paid to the potential for positive responses towards depressed individuals, especially from individuals who normally receive substantial benefits from the depressed individual.

>>>>

Due to the ease in which signals can be faked [@Zahavi1975; @mokkonen2016] and the costs of accepting a deceptive signal as reliable [@dawkins1978; @johnstone1993; @Hasson1994], demonstrating that one’s need is genuine has likely been a recurrent adaptive problem since humans evolved the capacity to provide instrumental support to one another. One possible solution to this is to produce signals that are unprofitable to fake due to the costs involved. [@Zahavi1975]. This is possible whenever signaling is comparatively cheaper for those who have the relevant underlying quality relative to those who lack it [@grafen1990]. In these instances, individuals with the quality maximize their fitness at greater levels of investment in the signal than those who lack the quality or possess it to a lesser degree [@zahavi1977]. The end result of this is expected to be a signal that either tends to not be worth faking or is impossible to fake to a large degree due to selection against fakers constraining the physiology behind the signal [@zahavi1977; @gangestad2005].

@@

Distinguishing between the information provided by signals versus cues is a major challenge for signaling approaches to behavior. A signal differs from a cue in that it is intentionally sent by an organism to alter the perceptions and behaviors of others, while a cue is any piece of information that influences an individual's behavior or physiological state that was not created for communicative effect [@MaynardSmithHarper1995]. Regarding the signals of interest, suicide attempts are more likely to take the form of pure, external signals than other symptoms of depression. This is because low mood's impact on behavior for non-signaling benefits likely conveys information as cues in addition to what might exist for signaling purposes. In other words, the signaling component of depression may come from the upregulation of symptoms beyond what would be optimal for non-signaling purposes rather than the creation of entirely new symptoms. Suicidality is likely an exception to this, due to the likelihood most suicide attempts likely do not have non-signaling functions [see @deCatanzaro1991 for a possible exception].

@@

Consistent with costly signaling theory, costlier signals resulted in greater perceptions of need and increased the likelihood of participants helping the signaler than the less costly signals. This effect was largest in the "brother-in-law" and "romantic partner" vignettes, both of which involved claims of assault against participants' imagined daughters. Compared to these scenarios, the effect of signal type was much smaller in both the "basketball coach" and the "thwarted marriage" vignette. In the "basketball coach" vignette, part of the reason for this may have been the lack of relatedness between participants and the alleged victim. However, this was still a surprising finding due to the relatively small size of the request (i.e., recommend the coach be suspended rather than kick out family members or a romantic partner). With the "thwarted marriage" vignette, the fact that support came at the cost of one's younger daughter may explain the relatively small effect of signal strength. However, data from the Indian sample appears to be of lower quality, limiting our confidence in this interpretation (see limitations for more information). It is also possible that our research in designing the vignettes was insufficient for the population or the content of the vignette.

In the "brother-in-law" and "romantic partner" scenarios, *Crying* had little effect on the magnitude of pro-signaler responses relative to *Verbal request*, suggesting it was not costly enough to serve as a reliable signal in times of substantial conflicts of interests. In contrast, both *Depression* conditions increased support, albeit to similar degrees. One potential reason for this is the increase in symptoms from mild depression to depression was not costly enough to result in a large effect as the symptoms only differed in severity (e.g., grades dropping from As to Bs [mild depression] or Cs [depression]). Such small changes may be less impactful in vignette form than in real-life scenarios in which the effects of signaling may increase in severity and be prolonged over a lengthy period of time.

While the effect of *Suicide attempt* on belief at T2 was similar to the *Depression* conditions across vignettes, it resulted in greater likelihood of support. This may be due to our decision to measure belief in need as the degree participants believed the claims of the signaler (e.g., if the assaults actually happened) rather than a more inclusive measure. For this reason, signal cost may have provided additional information about the costliness of the signaler's adversity on top of information relevant to whether the event claimed actually happened.

Support for this interpretation comes from our mediation analyses, which showed that the likelihood of help was largely, but not entirely, mediated by signal's effect on belief in need in this study but totally mediated by belief in need when measured as belief in need of financial support in the pilot study when comparing both *Depression* and *Suicide attempt* (confirm suicide) to *Verbal request*. However, another possibility is that participants viewed the signal as providing information about adversity outside of the signaler's claims. For example, participants who saw the "brother-in-law" vignette may have thought the daughter was lying about the assault due to jealously over her cousin, with the stronger signals still conveying the severity of the competition between her and her cousin.

 However, the greater impact on likelihood of help than belief found in this study may have also arisen if participants viewed the signal as providing information about adversity outside of the signaler's claims.
## Relevance to costly signaling

@@

The need to explain suicide's connection to depression is often overlooked by explanations for major depression, particularly those which propose non-signaling functions. One explanation for their connection, is that both suicidality and depression may function to limit negative fitness outcomes in times of substantial adversity [@andrews2006; @Hagen2002; @Hagen2003; @HagenEtAl2008; @nock2008; @Rosenthal1993; @Stengel1956; @SymeEtAl2016], with costlier signals being required to elicit support when the payoffs to faking need are high enough so that less costly signals are not reliable indicators of one's need [@Rosenthal1993]. For this hypothesis to be correct, instrumental support would need to be provided frequently enough with large enough impacts so that the costs of signaling are outweighed by the benefits on average. In support of this, we found exposure to depression and suicide attempts resulted in greater belief in need and increased likelihood of support in fictional scenarios. This coupled with findings that depression can simultaneously elicit both positive and negative responses [@HokansonEtAl1986; @stephens1987], suggests that more attention should be paid to the potential for positive responses towards depressed individuals, especially from individuals who normally receive substantial benefits from the depressed or suicidal individual.
